\chapter{\bevezetes}

Positioned at the crossroads of physical simulation, and deep learning
techniques, our work is inspired by current advancements in physics-based deep
learning. We investigate the general problem of controlling simulation
parameters to achieve target outcomes.

\todo{"advacement" ki volt húzva kommenttel.}

Our novel method solves problems in fluid simulation by utilizing gradients of
a differentiable physics simulation. More specifically, we use gradients from
a reduced order fluid simulation technique based on eigenfunctions of the vector
Laplacian operator. Instead of simulating the full domain, a reduced dimensional
space is utilized, resulting in significant speed-ups. We show some of the
possibilities arising when differentiating a Laplacian eigenfluids simulation,
and investigate how the resulting gradients can drive optimization in different
scenarios.

\todo{The time of the simulation/optimization was actually never recorded,
    because it was so fast. (Though the time it took was actually logged to the
    console.) Every optimization epoch took a couple of seconds at most, and
    even the most involved ones ended in a couple of minutes (max 10-15 seconds
    for really long ones?)... It would be nice to insert some kind of benchmark
    somewhere.}

In this chapter, we briefly discuss previous research that served as inspiration
as well as a base for our current work. We also give an overview of the overall
structure of our thesis. 

\section{Previous Work}
\subsection{Fluid Simulation}
Most simulation methods are based on either an Eulerian (i.e.  grid-based), or
Lagrangian (i.e. particle-based) representation of the fluid.  For an overview
of fluid simulation techniques in computer graphics, see \cite{FluidNotes} and
\cite{BridsonFluid}. 

For advecting marker density in our fluid, as well as a comparative "baseline"
simulation, we will use Eulerian simulation techniques, mostly as described by
\cite{StableFluids}.

\subsubsection*{Reduced Order Modeling of Fluids}
Dimension reduction-based techniques have been applied to fluid simulation in
multiple previous works. \cite{Wiewel2019LatentSP} demonstrated that functions
of an evolving physics system can be predicted within the latent space of neural
networks. Their efficient encoder-decoder architecture predicted pressure
fields, yielding two orders of magnitudes faster simulation times than
a traditional pressure solver.

Recently, \cite{LatentSpaceSubdivision} predicted the
evolution of fluid flow via training a convolutional neural network (CNN) for
spatial compression, with another network predicting the temporal evolution in
this compressed subspace.  The main novelty of \cite{LatentSpaceSubdivision} was
the subdivision of the learned latent space, allowing interpretability, as well
as external control of quantities such as velocity and density of the fluid.

\subsubsection*{Eigenfluids}
Instead of \textit{learning} a reduced-order representation, another option is
to analytically derive the dimension reduction, and its time evolution.
\cite{dewitt} introduced a computationally efficient fluid simulation technique
to the computer graphics community. Rather than using an Eulerian grid or
Lagrangian particles, they represent fluid fields using a basis of global
functions defined over the entire simulation domain. The fluid velocity is
reconstructed as a linear combination of these bases.

They propose the use of Laplacian eigenfunctions as these global functions.
Following their method, the fluid simulation becomes a matter of evolving basis
coefficients in the space spanned by these eigenfunctions, resulting in
a speed-up characteristic of reduced-order methods. 

Following up on the work of \cite{dewitt}, multiple papers proposed improvements
to the use of Laplacian eigenfunctions for the simulation of incompressible
fluid flow. \cite{ModelReductionFluidSim} extended the technique to handle
arbitrarily-shaped domains. \cite{EigenfluidCompression} used Discrete Cosine
Transform (DCT) on the eigenfunctions for compression.
\cite{scalable-eigenfluids} improved scalability of the technique, and modified
the method to handle different types of boundary conditions.
\cite{scalable-eigenfluids} refers to the fluid simulation technique using
Laplacian eigenfunctions as \textit{eigenfluids}, which we also adhere to in the
following.

\subsection{Differentiable Solvers}
Differentiable solvers have shown tremendous success lately for optimization
problems, including training neural nework agents
(\cite{holl2019pdecontrol}, \cite{difftaichi}, \cite{warp2022}).

\cite{holl2019pdecontrol} address grid-based solvers, noting them as
particularly appropriate for dense, volumetric phenomena. They put forth
$\Phi_{Flow}$, an open-source simulation toolkit built for optimization and
machine learning applications, written mostly in Python.

\subsubsection*{Physics-based Deep Learning}
Despite being a topic of research for a long time (\cite{backprop}), the
interest in neural network algorithms is a relatively new phenomenon. This is
especially true for the use of learning-based methods in physical and numerical
simulations, which is a rapidly developing area of current research. Recently,
integrating physical solvers in such methods have been shown to outperform
previously used learning approaches (\cite{solver-in-the-loop}).

Drawing on a wide breadth of current research, \cite{pbdl} give an overview of
deep learning methods in the context of physical simulations. 

\section{Structure}
In chapter~\ref{sec:mathematicalFoundations}, we give an overview of
mathematical foundations, introducing notation used throughout the text, as well
as discuss  preliminaries we base later chapters on. 

In chapter~\ref{chapter:physical-simulations}, the basics of physical
simulations are introduced. Building on the multivariable calculus introduced in
chapter~\ref{sec:mathematicalFoundations}, the concept of differentiable physics
simulation is introduced. 

Diving deeper into a more specific area of physical simulations,
chapter~\ref{chapter:fluid-simulation} gives a deep dive into fluid simulation
techniques, discussing the Laplacian eigenfluids method more in-depth in
section~\ref{section:laplacian-eigenfluids}. 

Finally, as a culmination of all that came before,
chapter~\ref{chapter:controlling-laplacian-eigenfluids} describes our proposed
method of Controlling Laplacian Eigenfluids. 

We close our work with a short discussion and possible future directions in
chapter~\ref{chapter:discussion}.
